The interactions of rational pragmatic agents provide a framework for understanding efficient language structure and use

### 2019-01-17 paper/paper.Rmd
first changes from MF reading

abstract: 
    We discuss a framework for studying the distributional properties of linguistic systems as emerging from in-the-moment interactions of speakers and listeners. Our work takes Zipfian notions of lexicon-level efficiency as a starting point, connecting these ideas to Gricean notions of conversational-level efficiency. To do so, we begin by deriving an objective function for measuring the communicative efficiency of linguistic systems and then examining the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language.

@@

Why do natural languages look they way they do? While Zipf (1935) presented his “Principle of Least Effort” as a domain general framework for understanding human behavior, he took language as his central case-study. He proposed that distributional properties found in natural language were evidence of speaker-listener effort minimization. In his own words, “we are arguing that people do in fact act with a maximum economy of effort, and that therefore in the process of speaking-listening they will automatically minimize the expenditure of effort.” Evidence for this claim was largely derived at the level of the lexicon. Zipf argued that the particular relationship between a word’s frequency and its rank, length, and denotation size could be explained as an emergent property of speaker-listener effort minimization. \par

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows -- how does having \textbf{property x} make using language $\ell$ more or less useful for communication? This reframing provides an opportunity to study both language *structure* (as Zipf primarily did) or *use* in terms of efficiency. For example, Regier et al. (201X) showed that languages appear to organize semantic domains (form-meaning mappings) in maximally efficient ways. Likewise, Piantadosi et al. (2012) argued that lexical ambiguity is an efficient property of any communication system when communication is contextualized. We will return to Piantadosi’s argument later in the paper.\par

Zipf’s original work and subsequent functionalist projects describe a varied set of linguistic properties and human behaviors. Common to many is a characterization of a fundamental effort-asymmetry underlying everyday communication. Simply put, what is effortful as a speaker is likely different from what is effortful as a listener. Zipf characterized this as follows -- purely from the standpoint of speaker effort, what Zipf called "Speaker's Economy," an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of just a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. From the standpoint of listener effort, what Zipf called "Auditor's Economy," an optimal language $\ell_{listener}^*$ would bijectively map all possible meanings to distinct words, eliminating a listener's need to disambiguate. Here the asymmetry is operationalized by different costs -- speaker effort is related to *production cost* and listener effort to *understanding or disambiguation cost*. Clearly natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf proposed that the particular lexicon-level properties he observed emerged from the competition of these forces -- the pressure to jointly minimize speaker and listener effort.\par

This formulation begs the question -- at what level of analysis do local, in-the-moment interactions of speakers and listeners lead to optimized structure at the level of the lexicon? The example given by Zipf (1935) appears to describe local interaction in terms of a reference game. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance $m \rightarrow u$. Other functionalist projects have also assumed this basic reference game setting (Regier, et al. (201X); Piantadosi et al. (2012); Kirby, Smith & Folds (201X), OTHERS) and this simplification of the communicative act has proven productive in both theoretical (Ferrer-i-cancho, et al.,  201X), simulation-based (iterated learning stuff) and empirical explorations of efficient language structure and use (Some of Robert’s work here).

This move to study efficient language use in terms of reference games has clear theoretical underpinnings as well. A half century after Zipf, the linguist Lawrence Horn (1984), highlighted the importance of Zipf’s principles for explaining conversation-level phenomena. Horn suggested a direct link between Zipf’s Speaker and Listener economy and, what was at the time, more recent work on conversational pragmatics by Grice (1975). Horn highlighted that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom.” Put differently, system-level efficiency we see in languages is deeply related to local-level efficiency during the in-the-moment interactions of rational pragmatic speakers and listeners. \par

In this work, we present a step toward formalizing Horn’s observation -- connecting Zipfian notions of efficient language structure to Gricean notions of rational conversation. To do so minimally requires three basic ingredients -- (1) a language property we’d like to explain (2) a framework for describing in-the-moment interactions of speaker-listeners and (3) some measure of linguistic efficiency. For the current project we focus on the “communicative function of ambiguity” for our property -- translating the functionalist theory pursued by Piantadosi et al. (2012) into our framework. To model local, conversational interactions (2) we adopt the Rational Speech-act framework (RSA) (Frank & Goodman, 2012; Goodman & Frank, 2016). Finally, (3) we derive an objective function for measuring language efficiency in the reference game setting. Given this framework we can ask the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? This project is meant to demonstrate the ways in which a range of functionalist theories might make use of such a framework -- in theory replacing \textit{lexical ambiguity} with a range of behaviors and properties identified by functionalist approaches to language science.\par

@@

### Reference games
Zipf’s canonical (1935) example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that could be used to refer to a given meaning. Utterances may have different relative costs, operationalized via a prior over utterances $p(u)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $p(m)$. Note that the prior over meanings are analogous to the *need probabilities* assumed in previous work (Regier, CITATION). We consider a set of contexts $C$ which describe different need probability distributions over our set of meanings $p(m|c)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meanings pairs. The set $E$ all possible utterance-meaning mappings.

@@

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. This framework has proven productive in describing a range of phenomena in natural language understanding and generation including hyperbole, metaphor, scalar and ad-hoc implicature, among others (CITATIONS). RSA can largely be understood as a formalization of essential Gricean pragmatic principles -- interlocutors reasoning about one another and their context. For this reason, we adopt RSA as our representational framework to operationalize rational speaker-listeners. For a more extensive description of RSA as well as its application we refer readers to Goodman & Frank (2016), limiting the current discussion to essential, high-level details of the framework. \par

@@

## Optimal speaker and listener languages in the reference game setting
In terms of Zipf’s initial example about an optimal speaker or listener language we might consider just the first or second term of equation (1). That is, given our formulation we should expect optimal speaker languages $\ell_{speaker}*$ to minimize:

\begin{equation}
\ell_{speaker}* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(p(u))
\end{equation}

Likewise we might expect optimal listener languages $\ell_{listener}*$ to minimize:

\begin{equation}
\ell_{listener}* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(L(m|u;\ell))
\end{equation}

# Two case-studies on efficient language use and design
We now have a metric for exploring questions of language optimality in the reference game setting (the speaker-listener cross-entropy). In the following sections we examine the behavior of this objective with respect to the communicative function of ambiguity.

@@

In their paper, “The communicative function of ambiguity,” Piantadosi et al. (2012) essentially argue the opposite. They claim that ambiguity is an *efficient* property of any communication system in which *communication is contextualized*. We provide the heart of their argument here, however we refer to Piantadosi et al. (2012) for more detail:

*In an unambiguous linguistic system, $m_1$ is mapped to $\ell_1 \in L$ and $m_2$ is mapped to $\ell_2 \in L$, with $\ell_1 \neq \ell_2$. Suppose that $\ell_1$ is easier than $\ell_2$... If $\ell_1$ is easier than $\ell_2$, and $m_1$ and $m_2$ are well-disambiguated by context, then we can always create a linguistic system which is easier overall by mapping $m_1$ and $m_2$ both to $\ell_1$. This costs the same in terms of effort every time we communicate $m_1$, but saves effort every time we communicate $m_2$*\par

In words, the Piantadosi argument might be summarised as -- it is useful to have a language that re-uses low-cost material (has ambiguity) so long as the cost of disambiguating the material is low. They argue that disambiguation cost is low when context if informative.

@@

## Experiment 1:  Optimal languages contain ambiguity when context is informative

The argument outlined by Piantadosi et al. (2012) is clearly compatible with our current reference game setting with just a single addition -- context. The authors argue that mapping both $m_1$ and $m_2$ to $l_1$ is useful when they can disambiguated *in context*. In our case we consider a context $c$ to specify a re-ordering of the referent need probabilities. That is, $p(m|c_1) \neq p(m|c_2)$. In the two context case.

@@

\begin{equation}
\begin{split}
\sum_{c\in C}p(c)\sum_{u, m}S(u|m, c;\ell)p(m|c)-log[L(m|u,c;\ell)p(u)] = \\ \mathbb{E}_{c\sim P(c)[\mathbb{E}_{P_speaker(u, m, c; \ell)}[-log(P_listener(u, m, c; \ell))]}]
\end{split}
\end{equation}

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=3, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Left facet -- optimal languages are more likely to contain ambiguous items as amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal langauge under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration. Right facet -- Gains in efficiency increase with the amount of contextual information (as the number of contexts increases). Vertical axis shows the difference in cross-entropy when the number of contexts is 1 and when the number of contexts is greater than 1. Horizontal axis shows each of the four conditions (the four context sizes). Note that the difference for condition 1 is zero as there is only a single context."}
img1 <- grid::rasterGrob(as.raster(png::readPNG("figs/fig1.png")))
img2 <- grid::rasterGrob(as.raster(png::readPNG("figs/fig2.png")))
gridExtra::grid.arrange(grobs=list(img1, img2), ncol=2)
```

@@

#### Baselines
For comparison, we also examine properties of optimal languages under two  additional objectives. Zipf (1935) proposed that a language optimized for speaker-ease should consist of a single word. We operationalize this using the *first half* of equation (1).

\begin{equation}
\begin{split}
\text{Speaker-only objective} = \mathbb{E}_{u, m\sim P_{speaker}}[\text{speaker effort}] \\= \sum_{u, m}P_{speaker}(u, m; \ell)[-log(p(u))]
\end{split}
\end{equation}

Zipf also proposed that the optimal listener language should bijectively map all words to unique meanings. We operationalize this using the *second half* of equation (1).
\begin{equation}
\begin{split}
\text{Listener-only objective} = \mathbb{E}_{u, m\sim P_{speaker}}[\text{listener effort}] \\ = \sum_{u, m}P_{speaker}(u, m; \ell)[-log(L(m|u; \ell)]
\end{split}
\end{equation}

@@

## Experiment 2: Rational-pragmatic speakers use ambiguity efficiently

Our first experiment made a fairly direct test of the communicative function of ambiguity proposed theory presented by Piantadosi et al. (2012). Sampling a set of need probabilities and utterance costs we explored the space of possible languages, examining the properties of the language $\ell^*$ which minimized our objective. Results indicated that ambiguity is an efficient property when context is informative. This experiment, however, assumed that our speaker and listener agents could always disambiguate items context. That is, in our four conditions both agents had perfect knowledge of the conditional distributions $p(m|c_1), \dots, p(m|c_4)$. Consider the scenario in which a reader is beginning a news article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this scenario, using an ambiguous pronoun early in the article lead to misunderstanding. Using such ambiguity is only warranted when the appropriate person or event is salient in the context.\par

We consider such a scenario in experiment two -- analyzing efficient *use* of ambiguity over discourse. In this scenario, we assume a single language $\ell$ which contains ambiguous items, known to both the speaker and listener. The topic of conversation $t\in T$ defines the particular meaning distribution at hand -- $p(m|T=t)$. This conditional distribution is known only to the speaker a priori. Over the course of a discourse, $D$ the listener tries to infer both the current topic as well as the meanings. That is we consider a listener model $L(m, t|u,D;\ell)$ and speaker model $S(u|m,t,D;\ell)$. Note that both the speaker and listener can track the history of previous utterances $D$.\par

### Hypothesis
In this setting, if speakers are *using ambiguity efficiently* they should use the ambiguous material only when they know the listener can disambiguate the items in context. Since listeners accumulate information about the current topic distribution ($p(m|t)$) over as the discourse progresses, we should expect to see more frequent use of ambiguous items later in the discourse.

@@

We conduct $n=50$ simulations, generating discourses of length $|D|=50$.  We consider a single language $\ell$ with $|U|=6$ and $|M|=4$ specified by the boolean matrix:

@@

We assume that $p(u_5) = p(u_6) > p(u_{1,...,4})$. That is, the two ambiguous utterances are less costly than the non-ambiguous utterances. Over the discourse both agents have complete recall of the set of possible utterances by the speaker $D$. Note that the listener can attempt to infer the current topic distribution $p(C=c|D)$ via $p(C=c|D) \propto \prod_{i=0}^|D|S(u_i|m_i)p(m_i|c)$.

@@

The left facet of figure 3 shows the empirical probability that the speaker used an ambiguous utterance as a function of discourse position. The right facet shows the listener belief about the current topic $c \in C$ given the discourse history ($p(c|D)$). Note that the trajectory of these items are close -- the speaker is more likely to use ambiguous material only when the listener is able to disambiguate that material. Figure 4 shows the listener belief about the current context. Additionally we can track the behavior of our cross-entropy objective and it’s component speaker and listener elements over the discourse. Overall we see declines in the speaker-listener cross-entropy. These are derived primarily from the listener learning about the current context of conversation (i.e. the current $c$ defining $p(m|c)$), however we also see declines in speaker effort, as they are able to use loss-costly material later in the discourse.

@@

These demonstrated the efficient *use* of ambiguous material over a discourse, in line with the argument of Piantadosi (2012). While the speaker may prefer to use the less costly, ambiguous utterances at all points throughout the discourse, using the ambiguous items early leads to higher listener-level cost. Only once the context is known to the listener does it makes sense to use the lower-cost material.

>>>>

abstract: 
    We discuss a framework for studying how the distributional properties of linguistic systems emerge from in-the-moment interactions of speakers and listeners. Our work takes Zipfian notions of lexicon-level efficiency as a starting point, connecting these ideas to Gricean notions of conversational-level efficiency. To do so, we begin by deriving an objective function for measuring the communicative efficiency of linguistic systems and then examining the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems.

@@

Why do languages look they way they do? While Zipf (1935) proposed that distributional properties found in natural language were evidence of speaker-listener effort minimization. In his own words, “we are arguing that people do in fact act with a maximum economy of effort, and that therefore in the process of speaking-listening they will automatically minimize the expenditure of effort.” Evidence for this claim was largely derived at the level of the lexicon. Zipf argued that the particular relationship between a word’s frequency and its rank, length, and denotation size could be explained as an emergent property of speaker-listener effort minimization. \par

Zipf articulated what is now considered a *functionalist* approach to language science -- analyzing language structure and use in terms of efficiency. Such an approach might reframe our opening question as follows -- how does having \textbf{property x} make using language $\ell$ more or less useful for communication? This reframing provides an opportunity to study both language *structure* (as Zipf primarily did) or *use* in terms of efficiency. For example, Regier et al. (201X) showed that languages appear to organize semantic domains (form-meaning mappings) in maximally efficient ways. Likewise, Piantadosi et al. (2012) argued that lexical ambiguity is an efficient property of any communication system when communication is contextualized.

Zipf’s original work and subsequent functionalist projects describe a varied set of linguistic properties and human behaviors. Common to many is a characterization of a fundamental effort-asymmetry underlying everyday communication. Simply put, what is effortful for a speaker is likely different from what is effortful for a listener. Zipf characterized this as follows -- purely from the standpoint of speaker effort, an optimal language $\ell_{speaker}^*$ would tend toward a vocabulary of just a single, low-cost word. Given such a language, the full set of potential meanings would be conveyed using only that word, i.e. $\ell_{speaker}^*$ would be fully ambiguous and all possible meanings would need to be disambiguated by a listener. From the standpoint of listener effort, an optimal language $\ell_{listener}^*$ should map all possible meanings to distinct words, removing a listener's need to disambiguate. In this example, speaker effort is related to *production cost* and listener effort to *understanding or disambiguation cost*. Clearly natural languages fall between the two extremes of $\ell_{speaker}^*$ and $\ell_{listener}^*$. Zipf proposed that the particular lexicon-level properties he observed were a result of these competing forces -- the pressure to jointly minimize speaker and listener effort.\par

But at what level of does local, in-the-moment interaction lead to optimized structure at the level of the lexicon? The example given by Zipf (1935) appears to describe local interaction in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance $m \rightarrow u$. Other functionalist projects have also assumed this basic reference game setting (Regier, et al. (201X); Piantadosi et al. (2012); Kirby, Smith & Folds (201X), OTHERS) and this simplification of the communicative act has proven productive in both theoretical (Ferrer-i-cancho, et al.,  201X), simulation-based (iterated learning stuff) and empirical explorations of efficient language structure and use (Some of Robert’s work here).

This move to study efficient language use in terms of reference games has clear theoretical underpinnings as well. The linguist Lawrence Horn (1984), highlighted the importance of Zipf’s principles for explaining conversation-level phenomena. Horn suggested a direct link between Zipf’s Speaker and Listener economy and work on conversational pragmatics by Grice (1975). Horn highlighted that the interaction of Zipf’s forces were “largely responsible for generating Grice’s conversational maxims and the schema for pragmatic inference derived therefrom.” Put differently, system-level efficiency we see in languages is deeply related to local-level efficiency during the in-the-moment interactions of rational pragmatic speakers and listeners at the level of conversation. \par

In this work, we present a step toward formalizing the framing presented above  -- connecting Zipfian notions of efficient language structure to Gricean notions of rational conversation. To do so minimally requires three basic ingredients -- (1) a language property we’d like to explain, (2) a framework for describing in-the-moment interactions of speaker-listeners, and (3) some measure of linguistic efficiency. For the current project we focus on the “communicative function of ambiguity” for our property -- translating the functionalist theory pursued by Piantadosi et al. (2012) into our framework. To model local, conversational interactions (2) we adopt the Rational Speech-act framework (RSA) (Frank & Goodman, 2012; Goodman & Frank, 2016). Finally, (3) we derive an objective function for measuring language efficiency in the reference game setting. Given this framework we can ask the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? This project is meant to demonstrate the ways in which a range of functionalist theories might make use of such a framework -- in theory replacing \textit{lexical ambiguity} with a range of behaviors and properties identified by functionalist approaches to language science.\par

@@

### Reference games
Zipf’s canonical (1935) example of optimal speaker- and listener-languages took the form of a reference game (similar to those described by Wittgenstein). We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that could be used to refer to a given meaning. Utterances may have different relative costs, operationalized via a prior over utterances $p(u)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $p(m)$. Note that the prior over meanings are analogous to the *need probabilities* assumed in previous work (Regier, CITATION). We consider a set of contexts $C$ which describe different need probability distributions over our set of meanings $p(m|c)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meanings pairs. The set, $E$, of all possible utterance-meaning mappings.

@@

### Speakers and listeners
The Rational Speech-act framework (RSA) is computational-level theory of pragmatic language use. RSA can largely be understood as a formalization of essential Gricean pragmatic principles -- interlocutors reasoning about one another and their context. For this reason, we adopt RSA as our representational framework to operationalize rational speaker-listeners. For a more extensive description of RSA as well as its application we refer readers to Goodman & Frank (2016), limiting the current discussion to essential, high-level details of the framework. \par

@@

# Simulations

@@

In their paper, “The communicative function of ambiguity,” Piantadosi et al. (2012) essentially argue the opposite. They claim that ambiguity is an *efficient* property of any communication system in which *communication is contextualized*. Simply put, it is useful to have a language that re-uses low-cost material (has ambiguity) so long as the cost of disambiguating the material is low. They argue that disambiguation cost is low when context if informative.

@@

## Simulation 1:  Optimal languages contain ambiguity when context is informative

The argument outlined by Piantadosi et al. (2012) is clearly compatible with our current reference game setting with just a single addition -- context. The authors argue that mapping both $m_1$ and $m_2$ to $l_1$ is useful when they can disambiguated *in context*. In our case we consider a context $c$ to specify a re-ordering of the referent need probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.

@@

\begin{equation}
\begin{split}
\sum_{c\in C}p(c)\sum_{u, m}S(u|m, c;\ell)p(m|c)-log[L(m|u,c;\ell)p(u)] = \\ \mathbb{E}_{c\sim P(c)}[\mathbb{E}_{P_{speaker}(u, m, c; \ell)}[-log(P_{listener}(u, m, c; \ell))]]
\end{split}
\end{equation}

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=6, fig.height=3, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Left facet -- optimal languages are more likely to contain ambiguous items as amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal langauge under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration. Right facet -- Gains in efficiency increase with the amount of contextual information (as the number of contexts increases). Vertical axis shows the difference in cross-entropy when the number of contexts is 1 and when the number of contexts is greater than 1. Horizontal axis shows each of the four conditions (the four context sizes). Note that the difference for condition 1 is zero as there is only a single context."}

img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

#### Baselines
For comparison, we also examine properties of optimal languages under two  additional objectives. Zipf (1935) proposed that the optimal speaker language should consist of a single low cost word $\ell_{speaker}^*$. We operationalize this using the *first half* of equation (1).

\begin{equation}
\ell_{speaker}^* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(p(u))
\end{equation}

The optimal listener language $\ell_{listener}^*$, by contrast, should bijectively map all words to unique meanings. We operationalize this using the *second half* of equation (1).

\begin{equation}
\ell_{listener}^* = argmin_{\ell\in L}\mathbb{E}_{P_{speaker}(u, m; \ell)}(L(m|u;\ell))
\end{equation}

@@

## Simulation 2: Rational-pragmatic speakers use ambiguity efficiently

Our first experiment made a fairly direct test of the communicative function of ambiguity proposed theory presented by Piantadosi et al. (2012). Sampling a set of need probabilities and utterance costs, we explored the space of possible languages, examining the properties of the language $\ell^*$ which minimized our objective. Results indicated that ambiguity is an efficient property when context is informative. This simulation, however, assumed that our speaker and listener agents could always disambiguate items context. That is, in our four conditions both agents had perfect knowledge of the conditional distributions $p(m|c_1), \dots, p(m|c_4)$. Consider the scenario in which a reader is beginning a news article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this scenario, using an ambiguous pronoun early in the article lead to misunderstanding. Using such ambiguity is only warranted when the appropriate person or event is salient in the context.\par

We consider such a scenario in simulation two -- analyzing efficient *use* of ambiguity over discourse. In this scenario, we assume a single language $\ell$ which contains ambiguous items, known to both the speaker and listener. Crucially, we do not assume that a listener has access the conditional distributions $p(m|c)$. Rather, we assume that the listener has knowledge of the set of possible context $C = \{c_1, \dots, c_k\}$, but does not know which one is currently being used by the speaker. In this setting, one can thing of the “contexts” as analogous to “topics” of conversation, where a “topic” specifies a particular seed of need probabilities $p(m|c)$.\par

Over the course of a discourse, $D$ the listener tries to infer both the current topic, $c$, as well as the particular meaning of a given utterance. That is we consider a listener model $L(m, c|u,D;\ell)$ and speaker model $S(u|m,c,D;\ell)$. Note that both the speaker and listener can track the history of previous utterances $D$.\par

### Hypothesis
In this setting, if speakers are *using ambiguity efficiently* they should only use the ambiguous material when they know the listener can disambiguate the items in context. Since listeners accumulate information about the current topic distribution ($p(m|t)$) over as the discourse progresses, we should expect to see more frequent use of ambiguous items later in the discourse.

@@

We conduct $n=200$ simulations, generating discourses of length $|D|=30$ utterances.  We consider a single language $\ell$ with $|U|=6$ and $|M|=4$ specified by the boolean matrix:

@@

We assume that $p(u_5) = p(u_6) > p(u_{1,...,4})$. That is, the two ambiguous utterances are less costly than the non-ambiguous utterances. In this setting, it is less costly for the speaker to use the ambiguous utterances. However, if the listener cannot disambiguate them, then they will incur a disambiguation cose. We assume that over the discourse both agents have complete recall of the set of possible utterances by the speaker $D$. This means that the listener can infer the current topic distribution $p(C=c|D)$ via $p(C=c|D) \propto \prod_{i=0}^|D|S(u_i|m_i)p(m_i|c)$. In this way, an efficient speaker should should avoid ambiguity \textit{early} in the discourse when the listener is unsure of the context, only using it later once they are confident the listener knows the context $p(m|C=c)$.

@@

The left facet of figure 3 shows the empirical probability that an RSA speaker used an ambiguous utterance as a function of discourse position. The right facet shows the listener belief about the current topic $c \in C$ given the discourse history ($p(c|D)$). Note that the trajectory of these items are close -- the speaker is more likely to use ambiguous material only when the listener is able to disambiguate that material. Figure 4 shows the listener belief about the current context. Additionally, we can track the behavior of our cross-entropy objective and it’s component speaker and listener elements over the discourse. Overall we see declines in the speaker-listener cross-entropy. These are derived primarily from the listener learning about the current context of conversation (i.e. the current $c$ defining $p(m|c)$), however we also see declines in speaker effort, as they are able to use loss-costly material later in the discourse.

@@

These simulations demonstrated the efficient *use* of ambiguous material over a discourse, in an alternative instantiation of the argument outlined by Piantadosi et al. (2012) -- Only once the context is known to the listener does it makes sense to use the lower-cost material.
